---
title: "Mandatory Assigment 1"
author: "Jone Damås - 256140"
date: "2022-09-14"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
fig.align = 'center'

```

### Exercise 1:

#### a.)

The function is a probability density function if the integral over the whole distribution equals 1:

$$
\int_{-\infty}^{\infty} f(x) \,dx=1
$$

$$
\int_{0}^{2} \frac{x}{2} \,dx=\frac{x^2}{4}\Big|_0^2=\frac{2^2}{4}-\frac{0^2}{4}=1
$$

Expected value is calculated by:

$$
E(X)=\int_{-\infty}^{\infty} x\cdot f(x) \,dx
$$

$$
\int_{0}^{2} \frac{x^2}{2} \,dx=\frac{x^3}{6}\Big|_0^2=\frac{2^3}{6}-\frac{0^3}{6}=\frac{4}{3}=1.333
$$

Variance is calculated by:

$$
Var(X)=E(X^2)-E(X)^2
$$

$$
E(X^2)=\int_{0}^{2} \frac{x^3}{2} \,dx=\frac{x^4}{8}\Big|_0^2=\frac{2^4}{8}-\frac{0^4}{8}=2
$$

$$
Var(X)=2-\Big( \frac{4}{3} \Big)^2=\frac{2}{9}=0.222
$$

$$
E\Big(\frac{1}{2}(X+1)\Big)=\int_{-\infty}^{\infty} \frac{1}{2}(x+1)\cdot f(x) \,dx
$$

$$
\frac{1}{2}\int_{0}^{2} \frac{x^2}{2} \,dx+\frac{1}{2}\int_{0}^{2} \frac{x}{2} \,dx=\frac{x^3}{12}\Big|_0^2+\frac{x^2}{8}\Big|_0^2=\frac{2^3}{12}+\frac{2^2}{8}=\frac{7}{6}=1.166
$$

$$
E(X^3)=\int_{-\infty}^{\infty} x^3\cdot f(x) \,dx
$$

$$
\int_{0}^{2} \frac{x^4}{2} \,dx=\frac{x^5}{10}\Big|_0^2=\frac{2^5}{10}-\frac{0^5}{10}=\frac{16}{5}=3.200
$$

#### b.)

The cumulative distribution function is found by:

$$
F(x)=\int_{-\infty}^{x} f(t) \,dt=\int_{0}^{x} \frac{t}{2} \,dx=\frac{t^2}{4}\Big|_0^x=\frac{x^2}{4}
$$

Using the cumulative distribution function:

$$
P(X<1)=F(1)=\frac{1^2}{4}=\frac{1}{4}=0.250
$$

$$
P(0.5<X<1.5)=F(1.5)-F(0.5)=\frac{1.5^2}{4}-\frac{0.5^2}{4}=\frac{1}{2}=0.500
$$

Finding the inverse cumulative distribution function:

$$
u=\frac{x^2}{4}\Rightarrow x=\pm \sqrt{4u}\Rightarrow x= \sqrt{4u}
$$

Only positive values for x gives positive probabilities, therefore, only the square root of 4 times u makes sense to use.

#### c.)

```{r}
rm(list=ls())

# Function returning n random numbers with density f(x) by the inverse transform method:
f_pdf_IT <- function(n){

  u <- runif(n, 0, 1)
  return(sqrt(4 * u))
}
# Start time measuring:
start_time <- Sys.time()

# Histogram of the 100000 random numbers with density f(x):
hist(f_pdf_IT(100000), probability=TRUE, ylim=c(0, 1), xlab='x', main='Histogram of f(x), Inverse Transform')

# Stop time measuring:
end_time <- Sys.time()

cat('Duration:', end_time - start_time)

```

#### d.)

Setting g(x)=1/2. As g(x) is constant, f/g is then proportional to f, and the function is maximized at the higher endpoint at x=2. Therefore:

$$
c=\frac{f(2)}{g(2)}=\frac{\frac{2}{2}}{\frac{1}{2}}=2
$$

#### e.)

```{r}
# Function returning n random numbers with density f(x) by the accept-reject method:

# c value obtained from calculations:
c <- 2

# Draw dunction
AR_draw <- function(){
  while(TRUE){
    x <- runif(1, min=0, max=2)
    a <- (x/2)/(c * 1/2)
    if(runif(1) < a) return(x)
  }
}

# Function which generates a pdf with the draw function:
f_pdf_AR <- function(n){
  v <- vector(length=n)
  for(i in 1:n){
    v[i] <- AR_draw()
  }
  return(v)
}

# Start measurement of time
start_time <- Sys.time()

# Histogram of the pdf:
hist(f_pdf_AR(100000), probability=TRUE, ylim=c(0,1), xlab='x', main='Histogram of f(x), Accept-reject')

# Stopmeasurement of time
end_time <- Sys.time()

cat('Duration:', end_time - start_time)

```

The inverse transform algorithm used 0.0391 seconds and the accept-reject algorithm used 1.243 seconds to execute the calculations and plot. This clearly demonstrates that the inverse transform algorithm is superior in terms of calculation speed.

#### f.)

Checking expectations and variance for the distributions created by the inverse transform and accept-reject methods:

```{r}

# Create matrix for result storage
data <- matrix(ncol=2, nrow=2)
# Expected value of the random numbers created from the inverse transform method:
data[1, 1] <- mean(f_pdf_IT(100000))
# Expected value of the random numbers created from the accept-reject method:
data[2, 1] <- mean(f_pdf_AR(100000))

# Variance of the random numbers created from the inverse transform method:
data[1, 2] <- var(f_pdf_IT(100000))

# Variance of the random numbers created from the accept-reject method:
data[2, 2] <- var(f_pdf_AR(100000))

# Setting row and column names for result table:
colnames(data) = c('E(X)','Var(X)')
rownames(data) <- c('Inverse Transform','Accept-reject')

# Creating result table:
tbl <- as.table(data)
print(tbl)


```

#### g.)

```{r}
# Kernel density estimation:
plot(density(f_pdf_IT(100000)), xlim=c(-0.5, 2.5), ylim=c(0, 1), xlab='x', main='Kernel density plot vs true density')

# pdf function:
f <- function(x){
  return(x/2)
}

# Creating true probability density from function:
x0 <- seq(-1, 3, 0.001)
f_pdf <- rep(0, times=length(x0))
f_pdf[x0>=0 & x0<=2] <- f(x0[x0>=0 & x0<=2])
lines(x0, f_pdf, col='red')

legend("topleft", legend = c("Kernel density", "True density"), lwd = 2, col = c("black", "red"))


```

### Exercise 2:

#### a.)

$$
E(A)=E \Big(\sum_{i=1}^{n}{X_i}\Big)=n \cdot E(X_i)=n \mu_i=n
$$

$$
Var(A)=Var\Big(\sum_{i=1}^{n}{X_i}\Big)=n\cdot Var(X_i)=n\cdot \sigma^2=n
$$

$$
E(B)=E \Big(\frac{1}{n}\sum_{i=1}^{n}{X_i}\Big)=\frac{1}{n}(n\cdot\mu_i)=\mu_i=1
$$

$$
Var(B)=Var\Big(\frac{1}{n}\sum_{i=1}^{n}{X_i}\Big)=\frac{1}{n^2}\cdot n\cdot Var(X_i)=\frac{\sigma^2}{n}=\frac{1}{n}
$$

C has a normal distribution with \~N(0,1) when n is large, and is explained by the central limit theorem.$$
\sqrt{n}\Big(\frac{1}{n}\sum_{i=1}^{n}{X_i}-1\Big)
$$

#### b.)

```{r}

# Number of values:
n <- 1000
# Number of simulation:
Nsim <- 100


A_vec <- vector(length=Nsim)
B_vec <- vector(length=Nsim)
C_vec <- vector(length=Nsim)

# Data generation for distribution A, B and C:
for(i in 1:100){
  A_vec[i] <- sum(rexp(n))
  B_vec[i] <- mean(rexp(n))
  C_vec[i] <- sqrt(n)*(mean(rexp(n))-1)
}

# Histogram of distributions for A,B and C:
hist(A_vec, prob=T, nclass=20, main='Histogram of A', xlab='x')
hist(B_vec, prob=T, nclass=20, main='Histogram of B', xlab='x')
hist(C_vec, prob=T, nclass=20, main='Histogram of C', xlab='x')

```

#### c.)

```{r}

n_list <- c(10, 100, 1000)

data <- matrix(nrow=length(n_list), ncol=2)

# Confidece interval generation:
for(j in 1:length(n_list)){
  B_vec <- vector(length=length(n_list))
  n <- n_list[j]
  
  for(i in 1:n){
    B_vec[i] <- mean(rexp(n_list))
  }
  B <- mean(B_vec)
  quant <- 1.96 * sd(B_vec)/sqrt(length(B_vec))
  data[j ,1] <- B - quant
  data[j ,2] <- B + quant
}

colnames(data) = c('lower', 'upper')
rownames(data) <- n_list

# Creating result table:
tbl <- as.table(data)
print(tbl)

```

### Exercise 3:

#### a.)

Considering the complexity of finding the inverse of the cumulative distribution function of f(x), the best approach would be to use the accept-reject method.

Setting g(x)=1/2. As g(x) is constant, f/g is then proportional to f, and the function is maximized at the lower endpoint at x=0. Therefore:

$$
c=\frac{f(0)}{g(0)}=\frac{\frac{1}{3}(e^{-0}+e^{-\frac{0}{2}})}{\frac{1}{2}}=\frac{4}{3}
$$

#### b.)

```{r 3}

fx <- function(x){
  return(1/3*(exp(-x)+exp(-x/2)))
}

# The constant c:
c <- 4/3

fx_AR_draw <- function(){
  while(TRUE){
    x <- runif(1, min=0, max=12)
    a <- fx(x)/(c * 1/2)
    if(runif(1) < a) return(x)
  }
}

fx_pdf_AR <- function(n){
  v <- vector(length=n)
  for(i in 1:n){
    v[i] <- fx_AR_draw()
  }
  return(v)
}



# Adding true density plot:
hist(fx_pdf_AR(100000), prob=T, main='Histogram of f(x)', xlab= 'x', nclass=50)
curve(fx, col='red', type='l', lwd=2.0, xlim=c(0, 12), add=T)

```

### Exercise 4:

#### a.)

```{r}
# Load the lawn.R script:
source('lawn.R')

sim_Etime <- function(e, n){
  result <- vector(length = n)
  for(i in 1:n){
    result[i] <- sim.lawn(e)
  }
  return(result)
}

res <- sim_Etime(5, 200)
hist(res, prob=T, nclass=50)

cat('Mean:', mean(res), '\n')
cat('Median:', median(res), '\n')
cat('Standard deviation:', sd(res), '\n')

```

#### b.)

```{r}

# sequence of Etime values
Etime <- c(1.0, 5.0, 10.0, 100.0)
# Etime dataframe and table column names
names <- c('1', '5', '10', '100')

data <- matrix(nrow=length(Etime), ncol = 3)

for(i in 1:length(Etime)){
  
  result <- sim_Etime(Etime[i], 200)
  
  data[i, 1] <- mean(result)
  data[i, 2] <- median(result)
  data[i, 3] <- quantile(result, probs=0.05)
  
  if(i == 1){
    df <- data.frame('1'=result, check.names = FALSE)
  }
  else{
    df[names[i]] <- result
  }
}

colnames(data) = c('Mean','Median', '0.05-Quantile')
rownames(data) <- names

# Create table
tbl <- as.table(data)
print(tbl)

boxplot(df, col="orange", border="brown", main='Boxplot of simulated lawnmawer data', xlab='Avarage direction changes per second', ylab='Fraction of cells visited')


```

Based on the Boxplot results, the best choice is arguably the Etime=10 setting because of the smaller standard deviation, even though it has a lower fraction of the cells visited compared to Etime=100. The also Etime=100 has some lone points with a considerably lower fraction than that of Etime=10.
