---
title: "Mandatory Assigment 3"
author: "Jone Damås - 256140"
date: "2022-10-19"
header-includes:
   - \usepackage{bbm}
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
fig.align = 'center'
```

### Exercise 1:

#### a.)

$$
A=\int_{0}^{\infty}\exp(-x)\sqrt{x}\cdot dx=\int_{0}^{\infty}\sqrt{x}\cdot f(x)\cdot dx
$$

$$
\text{where}\quad  f(x)=exp(-1)\quad\Rightarrow\lambda=1
$$ $$
\hat{\theta_A}=\frac{1}{n}\sum_{i=1}^{n}\sqrt{X_i}\quad\text{where}\quad X\sim \exp(1)
$$

#### b.)

$$
B=\int_{0}^{10}\exp(-x)\sqrt{x}\cdot dx=\int_{0}^{\infty}I(x<10)\sqrt{x}\cdot f(x)\cdot dx
$$

$$
\text{where}\quad  f(x)=exp(-1)\quad\Rightarrow\lambda=1
$$ $$
\hat{\theta}_B=\frac{1}{n}\sum_{i=1}^{n}I(X_i<10)\sqrt{X_i}\quad\text{where}\quad X\sim \exp(1)
$$

#### c.)

$$
\hat{\theta}_C=\frac{2}{n}\sum_{i=1}^{n}\sqrt{1+\cos(X_i)}\quad\text{where}\quad X\sim \text{Uniform}(-1,1)
$$

$$
\hat{\theta_D}=\frac{2}{n}\sum_{i=1}^{n}\sqrt{1+\sin(X_i)}\quad\text{where}\quad X\sim \text{Uniform}(-1,1)
$$

#### d.)

$$
f(x)=\frac{1}{1^{\frac{3}{2}}\Gamma(\frac{3}{2})}x^{\frac{3}{2}-1}e^{-\frac{x}{1}}=\frac{2}{\sqrt{\pi}}\sqrt{x}\cdot e^{-x}
$$

$$
\hat{\theta}_{IS,A}=\frac{1}{n}\sum_{i=1}^{n}\frac{g(x)}{f(x)}=\frac{1}{n}\sum_{i=1}^{n}\frac{\sqrt{X_i}\cdot e^{-X_i}}{\frac{2}{\sqrt{\pi}}\sqrt{X_i}\cdot e^{-X_i}}=\frac{1}{n}\sum_{i=1}^{n}\frac{\sqrt{\pi}}{2}
$$

$$
\hat{\theta}_{IS,B}=\frac{1}{n}\sum_{i=1}^{n}\frac{I(x<10)g(x)}{f(x)}=\frac{1}{n}\sum_{i=1}^{n}\frac{I(X_i<10)\sqrt{X_i}\cdot e^{-X_i}}{\frac{2}{\sqrt{\pi}}\sqrt{X_i}\cdot e^{-X_i}}=\frac{1}{n}\sum_{i=1}^{n}I(X_i<10)\frac{\sqrt{\pi}}{2}
$$

The estimator ends up as a constant and gives the exact solution of the integral in the case of integral A, and a very good approximation for integral B. Therefore, this method is a good method for estimating A and B.

#### e.)

$$
f(x)=\frac{1}{(\frac{1}{4})^4\Gamma(4)}x^{4-1}e^{-4x}=\frac{128}{3}x^3\cdot e^{-4x}
$$

$$
\hat{\theta}_{IS,A}=\frac{1}{n}\sum_{i=1}^{n}\frac{g(x)}{f(x)}=\frac{1}{n}\sum_{i=1}^{n}\frac{\sqrt{X_i}\cdot e^{-X_i}}{\frac{128}{3}X_i^3\cdot e^{-4X_i}}=\frac{1}{n}\sum_{i=1}^{n}\frac{3}{128}X_i^{-\frac{5}{2}}\cdot e^{3X_i}
$$

$$
\hat{\theta}_{IS,B}=\frac{1}{n}\sum_{i=1}^{n}\frac{I(x<10)g(x)}{f(x)}=\frac{1}{n}\sum_{i=1}^{n}\frac{I(X_i<10)\sqrt{X_i}\cdot e^{-X_i}}{\frac{128}{3}X_i^3\cdot e^{-4X_i}}=\frac{1}{n}\sum_{i=1}^{n}I(X_i<10)\frac{3}{128}X_i^{-\frac{5}{2}}\cdot e^{3X_i}
$$

We see that f(x) is not close to g(x) in shape, which does not make for a flat g(x)/f(x). This is not great for reducing variance.

#### f.)

$$
V_i=a+b-U_i=-1+1-U_i=-U_i
$$

$$
\hat{\theta}_{AT,C}=\frac{b-a}{n}\Bigg(\sum_{i=1}^{n/2}g(U_i)+\sum_{i=1}^{n/2}g(V_i)\Bigg)=\frac{2}{n}\Bigg(\sum_{i=1}^{n/2}\sqrt{1+\cos(U_i)}+\sum_{i=1}^{n/2}\sqrt{1+\cos(-U_i)}\Bigg)
$$

$$
V_i=a+b-U_i=-1+1-U_i=-U_i
$$

$$
\hat{\theta}_{AT,D}=\frac{b-a}{n}\Bigg(\sum_{i=1}^{n/2}g(U_i)+\sum_{i=1}^{n/2}g(V_i)\Bigg)=\frac{2}{n}\Bigg(\sum_{i=1}^{n/2}\sqrt{1+\sin(U_i)}+\sum_{i=1}^{n/2}\sqrt{1+\sin(-U_i)}\Bigg)
$$

#### g.)

```{r}

# estimators for d)
est_A_d <- mean(rep(sqrt(pi)/2, 100000))

est_B_d <- mean((rgamma(100000, shape=3/2, rate=1) <= 10)*sqrt(pi)/2)

# estimators for e)
x <- rgamma(100000, shape=4, rate=1/4)
est_A_e <- mean(3/128*x^(-5/2)*exp(3*x))

x <- rgamma(100000, shape=4, rate=1/4)
est_B_e <- mean((x <= 10)*3/128*x^(-5/2)*exp(3*x))

# estimators for f)

# antithetic Monte Carlo:
u <- runif(100000/2, -1, 1)
v <- -u

est_C <- 2*mean(sqrt(1+cos(c(u, v))))

est_D <- 2*mean(sqrt(1+sin(c(u, v))))

print(paste('Estimaton of integral A with gamma(3/2,1):', round(est_A_d, 8)), quote=F)
print(paste('Estimaton of integral B with gamma(3/2,1):', round(est_B_d, 8)), quote=F)
print(paste('Estimaton of integral A with gamma(4,1/4):', round(est_A_e, 8)), quote=F)
print(paste('Estimaton of integral B with gamma(4,1/4):', round(est_B_e, 8)), quote=F)
print(paste('Estimaton of integral C:', round(est_C, 8)), quote=F)
print(paste('Estimaton of integral D:', round(est_D, 8)), quote=F)


```

As expected, the estimations of integral A and B using the gamma(3/2, 1) importance sampling method gave very accurate estimations. Its however apparent that the estimates for integral A and B using the gamma(4, 1/4) importance sampling method is not good at all. The antithetic Monte Carlo estimators work well for the estimations of integral C and D as expected by theory.

### Exercise 2:

#### a.)

$$
X\sim binomial(n,V_c(2x)^{-d})
$$

Because:

1.) We have a fixed number of random variables generated, n.

2.) Each outcome of $\textbf1(u_j)$ is independent.

3.) $\textbf1(u_j)$ has only two outcomes, either 0 or 1.

4.) The probability of the outcome is the same for each value.

The probability of success in a binomial distribution is defined by:

$$
p=\frac{E(X)}{n}
$$

We see that $E(X)$ is $X$ in our case, since $X=\sum_{i=1}^{n}\textbf1(u_j)$.

$p$ is therefore:

$$
\hat{V_c}=\frac{(2c)^d}{n}X \Rightarrow\frac{X}{n}=\hat{V_c}(2c)^{-d}\Rightarrow p=\hat{V_c}(2c)^{-d}
$$

#### b.)

$$
\frac{SD(\hat{V}_1)}{V_1}<r\Rightarrow\frac{\sqrt{Var(\hat{V}_1)}}{V_1}<r\Rightarrow\frac{\sqrt{Var((2)^dX/n)}}{V_1}<r\Rightarrow\frac{Var((2)^dX)}{V_1^2n}<r^2
$$

$$
\Rightarrow\frac{2^{2d}Var(X)}{V_1^2n}<r^2\Rightarrow n>\Bigg(\frac{2^{d}\sigma_X}{V_1r}\Bigg)^2
$$

Inserting expression for $V_1$:

$$
n>\Bigg(\frac{2^d\sigma_X}{(2^d/d!)r}\Bigg)^2\Rightarrow n>\Bigg(\frac{d!\cdot\sigma_X}{r}\Bigg)^2
$$

As d is in a factorial expression and squared for the minimum number of simulations. There would be needed a rapidly increasing number of simulations as the dimension (d) increases.

#### c.)

```{r}

# Estimation function:
est_V <- function(d, c, n){
  X <- 0
  for(j in 1:n){
    u <- runif(d, min=-c, max=c)
    if(sum(abs(u))<=c){
      X = X + 1
    }
  }
  V <- X*(2*c)^d/n
  return(V)
}

# True value function:
tv <- function(d){
  return(2^d/(factorial(d)))
}

data <- matrix(nrow=3, ncol=2)

data[1,1] <- est_V(2, 1, 100000)
data[2,1] <- est_V(4, 1, 100000)
data[3,1] <- est_V(8, 1, 100000)

data[1,2] <- tv(2)
data[2,2] <- tv(4)
data[3,2] <- tv(8)

colnames(data) <- c('Estimated values', 'True values:')
rownames(data) <- c('d=2:', 'd=4:', 'd=8:')

table <- as.table(data)
table
```

#### d.)

```{r}

V_c <- function(d, c, n, s){
  sum_vec <- vector(length=n)
  for(j in 1:n){
    z <- rnorm(d, mean = 0, sd=s)
    I <- 0
    # Indicator function on the vector z:
    if(sum(abs(z))<=c){
      I <- 1
    }
    product <- prod(dnorm(z, sd=s))
    sum_vec[j] <- I / product
  }
  V <- mean(sum_vec)
  SE <- sd(sum_vec)/tv(d)
  return(c(V, SE))
}

data <- matrix(nrow=3, ncol=2)

V_d2 <- V_c(2, 1, 100000, 1)
V_d4 <- V_c(4, 1, 100000, 1)
V_d8 <- V_c(8, 1, 100000, 0.2)

data[1,1] <- V_d2[1]
data[2,1] <- V_d4[1]
data[3,1] <- V_d8[1]

# Relative standard error
data[1,2] <- V_d2[2]
data[2,2] <- V_d4[2]
data[3,2] <- V_d8[2]

colnames(data) <- c('Estimated values', 'Relative standard error:')
rownames(data) <- c('d=2:', 'd=4:', 'd=8:')

table <- as.table(data)
table

```

#### e.)

```{r}
library(mvtnorm)

mu <- c(1/4, 1/4, 1/4, 1/4)

sig <- matrix(c(1, 1/4, 1/4, 1/4,
                1/4, 1, 1/4, 1/4,
                1/4, 1/4, 1, 1/4,
                1/4, 1/4, 1/4, 1), nrow=4)

a_sim <- function(n, c, mu, sigma){
  d <- length(mu)
  sum_vec <- vector(length=n)
    for(j in 1:n){
    
    x <- runif(4, -c, c)

    u <- dmvnorm(x, mean=mu, sigma=sigma)
    
    I <- 0
    if(sum(abs(x))<=c){
      I <- 1
    }
    sum_vec[j] <- u * I
  }
  return(c((2*c)^4*sum(sum_vec)/n, sd(sum_vec)))
}

print(paste('a=', round(a_sim(10000, 1, mu, sig)[1], 5)), quote=F)
print(paste('SD=', round(a_sim(10000, 1, mu, sig)[2], 5)), quote=F)

```

### Exercise 3:

#### a.)

```{r}

x <- read.table("M3data.txt")$V1

# KDE estimator
f1.hat <- function(x){
  return(density(x,from=1,to=1,n=1)$y)
}

f1.o <- f1.hat(x)
print(paste0("Estimate of f(1) : ",f1.o), quote=F)

B <- 2000

n <- length(x)
f1.estimate <- numeric(B)

for(i in 1:B){
  f1.estimate[i] <- f1.hat(sample(x,size=n,replace=TRUE))
}

hist(f1.estimate,prob=TRUE,nclass=sqrt(B)/2, main='Estimate of f(1)', xlab=expression(theta))  
abline(v=f1.hat(x),col="red",lwd=3)

# Standard error
print(paste('Standard error:', round(sd(f1.estimate), 6)), quote=F)
# Bias
print(paste('Bias:', round(mean(f1.estimate) - f1.o, 6)), quote=F)

```

#### b.)

```{r}

# Function for estimating f(1) using a histogram:
snr <- function(x){
  n <- length(x)
  w <- 3.49/2 * sd(x) * n^(-1/3)
  
  # Number of observation in interval:
  A <- sum(x > (1-w) & x < (1+w))
  
  # Scott's Normal Reference rule:
  estimator <- A / (2*n*w)
  return(estimator)
}

# Import data from file:
data <- read.table("M3data.txt")$V1
# Apply function to data:
print(paste('Estimator:', round(snr(data), 5)), quote=F)

```

#### c.)

```{r}

B <- 2000

n <- length(x)
f1.estimate <- numeric(B)

for(i in 1:B){
  f1.estimate[i] <- f1.hat(sample(x,size=n,replace=TRUE))
}

# Standard error
print(paste('Standard error:', round(sd(f1.estimate), 6)), quote=F)
# Bias
print(paste('Bias:', round(mean(f1.estimate) - snr(data), 6)), quote=F)


```

The alternative method using a histogram has a larger Bias than the KDE-based estimate of f(1), and would therefore be the preferred method for estimation.

### Exercise 4.)

#### a.)

```{r}

Nsim <- 10000

res_matrix <- matrix(nrow=Nsim, ncol=3)

for(i in 1:Nsim){
  n <- 32
  x <- numeric(n)
  x[1] <- 1.0
  x[2] <- 0.8
  
  for(j in 3:n){
    x[j] <- min(4.0,rexp(1)*(0.4+0.7*x[j-1]+0.1*x[j-2]))
    }

  # Finding the total production the next 30 days:
  res_matrix[i,1] <- sum(x[3:n])
  # Finding number of days at full capacity:
  res_matrix[i,2] <- length(x[x == 4.0])
  # Finding the minimum value of the simulation:
  res_matrix[i,3] <- min(x)
}

par(mfrow=c(2,2))

hist(res_matrix[,1], prob=T, main='Total production in the next 30 days', breaks=20, xlab='x')
hist(res_matrix[,2], prob=T, main='Days at full capacity', xlab='Days')
hist(res_matrix[,3], prob=T, main='Smallest daily output', breaks=30, xlab='x', xlim=c(0,0.4))

tp <- c(round(mean(res_matrix[,1]), 3), round(quantile(res_matrix[,1], c(.25, .50, .75), names = F), 2))
fc <- c(round(mean(res_matrix[,2]), 4), round(quantile(res_matrix[,2], c(.25, .50, .75), names = F), 2))
mn <- c(round(mean(res_matrix[,3]), 4), round(quantile(res_matrix[,3], c(.25, .50, .75), names = F), 4))

res_data <- rbind(tp, fc, mn)

colnames(res_data) <- c('Mean', 0.25, 0.50, 0.75)
rownames(res_data) <- c('Total production:', 'Days at full cap.:', 'Smallest daily output:')

table <- as.table(res_data)
table

```

#### c.)

```{r}
n <- length(res_matrix[,1])

S_vector <- vector(length=n)
for(i in 1:n){
  S_vector[i] <- max(c(0, 30 - res_matrix[i,1]))
}

price <- rgamma(n, shape=10, scale=1)
res <- S_vector * price

money <- quantile(res, 0.95, names=F)
print(paste('Money needed for 95% certanty:', round(money, 2)), quote=F)

```
